---
title: "Making Outlier Charts"
output: html_document
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Install packages

First install the necessary packages using install.packages("packagename"). [See names in libraries below or copy below]
install.packages("readr")
install.packages("tidyr")
install.packages("tidyverse")
install.packages("ggthemes")
install.packages("svglite")
install.packages('extrafont')


```{r libraries, echo = TRUE, results= FALSE, message = FALSE}
library(readr)
library(tidyr)
library(extrafont)
library(tidyverse)
library(ggthemes)
library(svglite)
```

### Import fonts
```{r fonts, echo = TRUE, results = FALSE, message = FALSE}
font_import()
```

You can view available fonts and their names by running: 
```{r fonts2, echo = TRUE, results = FALSE, message = FALSE }
fonts()
```

### Import the Outlier theme
You can import the Outlier styles in one of three ways: (1) Copy the code directly into the script (2) Import the styles from a file on your harddrive (3) Import the styles from GitHub (recommended).

#### 1 - Copy Outlier theme code into your script
```{r outlier1, }
theme_outlier <- function () { 
  font <- "Roboto Condensed"
  theme_fivethirtyeight(base_size = 12, base_family = font) %+replace%
    theme(
      plot.title = element_text(hjust = 0, vjust = 1.5, lineheight=.8, face="bold", size = 20, family=font, margin = margin(5,0,0,0)),
      plot.subtitle = element_text(hjust = 0, lineheight=.8, size = 15, family=font, margin = margin(0,0,20,0)),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      panel.grid.major.y = element_line( size=.5, color="gray", linetype = "dotted"), 
      axis.line = element_blank(),
      axis.ticks.x = element_line( size = .5, color = "gray", linetype="solid"),
      axis.text = element_text(size = 12),
      axis.text.x = element_text(vjust = 0),
      axis.ticks.length=unit(0.2, "cm"), 
      legend.text.align = 0,
      legend.title = element_blank(),
      legend.key = element_blank(),
      legend.text = element_text(family=font, size=10, color="#222222"),
      legend.position = c(0.25, .99),
      legend.direction = "horizontal",
      legend.background = element_rect(fill = "#f0f0f0", color = "lightgray", size = .2, linetype = "solid"),
      legend.key.width = unit(0.5,"cm"),
    )
}
```

#### 2 - Import Outlier styles from a file on your harddrive
```{r outlier2}
source("outlier_theme.R")
```

#### 3 - Import Outlier styles from GitHub (recommended)
```{r outlier3}
source("https://raw.githubusercontent.com/alastairotter/outlier-theme/main/outlier_theme.R")
```

### Making a bar chart
#### Import data from file
```{r data, echo = TRUE, results= FALSE, message = FALSE} 
vax <- read_csv("~/Documents/R-Projects/vaccinations/data-in/sa-vaccinations.csv")
```
#### View data
```{r dataview}
head(vax)
```

```{r dataview2}
summary(vax)
```

#### Set up bar chart
```{r setupbar}
barChart <- ggplot(vax, aes(date, vaccinated_daily)) + # start ggplot and set x and y
  geom_bar(stat = "identity", color="#f0f0f0", fill = "#70ADAD") + # set bar colors
  geom_hline(yintercept = 0, size = .5, color="gray") + # add horizontal line at 0 pioint for baseline
  labs(title="Vaccinations in South Africa",
       subtitle = "Daily vaccinations administered") + # set title and subtitle
  scale_y_continuous(labels = scales::comma) + # use commas as thousand separators
  coord_cartesian( ylim=c(-5,40000), expand = FALSE ) + # sets x-axis labels just below baseline
  theme_outlier() # sets outlier theme as the default style
```

#### Draw barchart plot
```{r plotbar}
barChart
```





#### Save barchart as an SVG file for Figma, or as a PNG
```{r barchartsvg}
ggsave(file="examples/barchart.svg", plot=barChart, width=10, height=5) # set file name
ggsave(file="examples/barchart.png", plot=barChart, width=10, height=5) # set file name
```


### Making a multi-line chart
#### Import data from file
```{r linedata, echo = TRUE, results= FALSE, message = FALSE}
vaxrec <- read_csv("~/Documents/R-Projects/vaccinations/data-in/cases-versus-recoveries.csv")
```
#### View data
```{r viewdata1}
head(vaxrec)
```
```{r viewdata2}
summary(vaxrec)
```
#### Use tidyverse to combine data fields
This is the preferred method although the same could be achieved with making two geom_line calls. 
```{r combinedata}
vaxrecnew <- vaxrec %>%
  select(date, cumulative_cases, cumulative_recoveries) %>%
  gather(key = "variable", value = "value", -date)
```
#### Set up line chart
```{r linesetup}
lineChart <- ggplot(vaxrecnew, aes(x = date, y = value)) + 
  geom_line(aes(color = variable)) + 
  scale_color_manual(labels = c("Total Cases", "Total Recoveries"), values = c("indianred", "#70ADAD")) + 
  scale_linetype_manual(values = c("solid", "dashed")) + 
  geom_hline(yintercept = 0, size = .5, color="gray") + 
  labs(title="Vaccinations in South Africa",
       subtitle = "Total vaccinations administered") + 
  scale_y_continuous(labels = scales::comma) + 
  coord_cartesian( ylim=c(-5,max(vaxrecnew$value)* 1.1), expand = FALSE ) + 
  theme_outlier() +
  theme(legend.position = c(0.22, .99))
```
#### Set up line chart (annotated)
```{r linesetupann}
#start ggplot and set x and y. Each line needs a "+" at the end if there is more code
lineChart <- ggplot(vaxrecnew, aes(x = date, y = value)) + 

# use the variable name as the color
geom_line(aes(color = variable)) + 

# set the labels and color for each line in the legend
scale_color_manual(labels = c("Total Cases", "Total Recoveries"), values = c("indianred", "#70ADAD")) +

# set the line type for each line
scale_linetype_manual(values = c("solid", "dashed")) + 

# add horizontal rule for baseline
geom_hline(yintercept = 0, size = .5, color="gray") + 

# sets title and sub-title
labs(title="Vaccinations in South Africa",
    subtitle = "Total vaccinations administered") + 

# use commas as thousand separators
scale_y_continuous(labels = scales::comma) + 

# sets x axis correctly
coord_cartesian( ylim=c(-5,max(vaxrecnew$value)* 1.1), expand = FALSE ) + 

# sets Outlier theme as the default style
theme_outlier() + 
  
# adjust position of legend
theme(legend.position = c(0.22, .99))
```

#### Options for charts
You can override styles in the Outlier theme with some of these below.
```{r options}
#theme(legend.position = c(0.22, .99))  # changes position of legend c(x,y)  
#theme(legend.background = element_blank())   removes legend background
#theme(legend.position = "none") removes legend
#theme(legend.position = "bottom")  puts legend below chart (other options: left, right, top)
```

#### Draw line chart
```{r drawlinechart}
lineChart
```
#### Save line chart as SVG for Figma
```{r linesvg}
ggsave(file="examples/line.svg", plot=lineChart, width=10, height=5)
ggsave(file="examples/line.png", plot=lineChart, width=10, height=5)
```


